---
title: "Identifying various forms of conserved genes using cliques (master file)"
author: "Ellen Dimmen Chapple"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: hide
    toc: true
    number_sections: true
    toc_float: true
    theme: spacelab
editor_options:
  chunk_output_type: console
---


```{r message=FALSE, warning=FALSE}

library(tidyverse)
library(igraph)

```

# Identifying cliques

Step 1 - removes orthogroups with no cliques. Format of input data frame is the same as for comparison files, output is a chr vector. Usefull for reducing runtime and as an overview.

Step 2 - clique code. Similar to step 1 but includes prefiltering steps such as discarding orthogroups that are too large and setting a maximum number of cliques to identify per orthogroup.Iterates through orthogroups generated in step 1, and  Saves the clique genes.

Step 3 - identify conserved and differentiated genes based on number of members within the pair groups (angiosperm, gymnosperm, cross-clade).

NOTE! Step 1 and 2 can be skipped if already run (and output hopefully saved).

**Step 1: remove orthogroups with no cliques**

INPUT FILE(S): the comparison files (COmPlEx) compiled into a single file (all_expressed_genes, see script: basic_files.R). 
OUTPUT FILE(S): CONS_OGS_RED which contains all orthogroups with at least one clique (3-6 members)

```{r eval=FALSE, include=FALSE}

load("Data/DATA/all_expressed_genes.RData")

expressologs <- expr_genes %>%
  filter(Max.p.Val < 0.9) %>%
  group_by(OrthoGroup) %>% 
  mutate(OG_size = n()) %>% 
  ungroup() %>% 
  arrange(desc(OG_size))

cat("Largest orthogroup: ", max(expressologs$OG_size), " gene pairs.", "\n", "Average orthogroup: ", round(mean(expressologs$OG_size), 0), " gene pairs.", "\n", "Median number of gene pairs per orthogroup: ", round(median(expressologs$OG_size), 0), ".")  

unique(expressologs$OG_size)[1:10]
unique(expressologs$OrthoGroup)[1:10]

discard_list <- c("OG0000000",  "OG0000018", "OG0000001", "OG0000007")
expressologs <- expressologs %>% 
  filter(!(OrthoGroup %in% discard_list))

cat("Largest orthogroup: ", max(expressologs$OG_size), " gene pairs.", "\n", "Average orthogroup: ", round(mean(expressologs$OG_size), 0), " gene pairs.", "\n", "Median number of gene pairs per orthogroup: ", round(median(expressologs$OG_size), 0), ".")  
   
load("Data/DATA/ones_and_zeros_all.RData")

# Add summary columns and remove all orthogroups that no expressed gene pairs.
df_with_sums <- ones_and_zeros_all %>%
  mutate(Angiosperms = rowSums(. [1:3])) %>%
  mutate(Cross = rowSums(. [4:12]))%>%
  mutate(Gymnosperms = rowSums(. [13:15])) %>%
  mutate(Conserved = rowSums(. [1:15]))%>%
  filter(Conserved > 0)

cons_ogs <- df_with_sums%>% 
  filter(rownames(df_with_sums) %in% expressologs$OrthoGroup) %>% 
  rownames()

# Loop over all orthogroups from above and count the number of max cliques that contain at least three members.
cliques_per_og <- list()

for (g in cons_ogs) {
  print(g)
  # g <- "OG0004847" # g <- "OG0018434"
  
  expressologs_g <- expressologs %>% 
    filter(OrthoGroup == g) 
  
  expressologs_g <- expressologs_g %>% 
    mutate(UGeneSpecies1 = paste0(expressologs_g$Species1, "-", expressologs_g$GeneSpecies1),
           UGeneSpecies2 = paste0(expressologs_g$Species2, "-", expressologs_g$GeneSpecies2))
  nodes <- data.frame(name = unique(c(expressologs_g$UGeneSpecies1, expressologs_g$UGeneSpecies2)))
  edges <- data.frame(from = expressologs_g$UGeneSpecies1, 
                      to = expressologs_g$UGeneSpecies2)
  net <- graph_from_data_frame(edges, directed = FALSE, vertices = nodes)
  
  largest_clique <- max_cliques(net, min = 3)
  
  df <- data.frame(Orthogroup = g,
                   numbCliques = length(largest_clique),
                   minSize = min(lengths(largest_clique)),
                   maxSize = max(lengths(largest_clique)))

  cliques_per_og<- append(cliques_per_og, list(df))
 
}

cliques_per_og_unlisted <- plyr::ldply(cliques_per_og)

# 2) Remove orthogroups with no cliques (min clique size = 3). The loop simply goes through all orthogroups and counts the number of cliques that contain at least three members. We therefore need to remove all orthogroups with zero cliques.
numb_of_cliques_per_og <-cliques_per_og_unlisted %>% 
  arrange(desc(numbCliques)) %>% 
  filter(numbCliques > 0) 

paste0("Number of orthologs with at least one clique (3 to 6 members): ", length(unique(numb_of_cliques_per_og$Orthogroup)))

# 3) Vectorise and save! 
cons_og_red <- numb_of_cliques_per_og$Orthogroup
save(cons_og_red, file = "Data/Data/conserved_orthogroups_reduced.RData")
save(numb_of_cliques_per_og, file = "Data/Data/number_of_cliques_per_OG.RData")

```


**Step 2: The clique code**
PRODUCES: clique_genes_filterable which can be used for identifying all clique types.

```{r eval=FALSE, include=FALSE}

# load("Data/Data/number_of_cliques_per_OG.RData") # overview of the number of cliques each orthogroup has.

# ------------- LOADING FILES AND SETTING PARAMETERS -----------------
load("Data/DATA/conserved_orthogroups_reduced.RData") # 
load("Data/DATA/all_expressed_genes.RData")


expressologs <- expr_genes %>%
  filter(Max.p.Val < 0.9)


# ------------- CLIQUE LOOP -----------------

expressologs_from_max_cliques <- list()
max_numb_of_cliques <- 5500

for (i in 1:length(cons_og_red)) {
   # i <- 1
  print(paste0(i, "/", length(cons_og_red)))
  
  g <- cons_og_red[i]
  # g <- "OG0004847"
  
  expressologs_g <- expressologs %>% 
    filter(OrthoGroup == g) 
  
  expressologs_g <- expressologs_g %>% 
    mutate(UGeneSpecies1 = paste0(expressologs_g$Species1, "-", expressologs_g$GeneSpecies1),
           UGeneSpecies2 = paste0(expressologs_g$Species2, "-", expressologs_g$GeneSpecies2))
  
  nodes <- data.frame(name = unique(c(expressologs_g$UGeneSpecies1, expressologs_g$UGeneSpecies2)))
  edges <- data.frame(from = expressologs_g$UGeneSpecies1, 
                      to = expressologs_g$UGeneSpecies2)
  net <- graph_from_data_frame(edges, directed = FALSE, vertices = nodes)
  largest_clique <- max_cliques(net, min = 3)
  
  
  if(length(largest_clique) > max_numb_of_cliques){
    largest_clique_red <- largest_clique[1:max_numb_of_cliques]
  }else{largest_clique_red <- largest_clique}
  
  
  for(c in 1:length(largest_clique_red)){

    if (c %% 500 == 0) {
      cat(c, "\n")
    }
    
    clique <- largest_clique[[c]]
    clique_name <- attr(clique, "names")
    
    clique_species <- str_split_fixed(clique_name, "-", n = 2)[,1]
    clique_genes <-  str_split_fixed(clique_name, "-", n = 2)[,2]
    
    expressologs_in_clique <- expressologs_g %>%
      filter(UGeneSpecies1 %in% clique_name & UGeneSpecies2 %in% clique_name) %>%
      select(c(1:5, 8)) %>% 
      mutate(cliqueID = paste0(g,"_" ,c))
    
    
    expressologs_from_max_cliques <- append(expressologs_from_max_cliques, list(expressologs_in_clique))
    
  }
  
}

clique_genes_unlisted <- plyr::ldply(expressologs_from_max_cliques)

# ------------- SAVING POINT -----------------
save(clique_genes_unlisted, file = "Data/DATA/clique_genes_unlisted.RData")


#
load("Data/DATA/clique_genes_unlisted.RData")

# test <- clique_genes_unlisted %>% 
#   group_by(cliqueID) %>% 
#   slice(1) %>% 
#   ungroup() %>% 
#   group_by(OrthoGroup) %>% 
#   mutate(cliquesInOG = n())
# 
# head(test)


clique_genes_unlisted_prefilter <- clique_genes_unlisted %>% 
  group_by(cliqueID) %>% 
  mutate(MaxpValNegLog10 = -log10(Max.p.Val)) %>% 
  mutate(NegLog10CliqueSum = sum(MaxpValNegLog10)) %>% 
  ungroup()


# unique(clique_genes_unlisted_prefilter$species_pair)

# ------------- SAVING POINT -----------------
save(clique_genes_unlisted_prefilter, file = "Data/DATA/clique_genes_prefiltered.RData")

# ------------- FORMATTING DATA FRAME -----------------
load("Data/DATA/clique_genes_prefiltered.RData")


weighted_max_cliques <- clique_genes_unlisted_prefilter %>% 
  mutate(species_pair = paste0(Species1, Species2))

speciesPairs <- tibble(SpeciesPair = weighted_max_cliques %>% distinct(species_pair) %>% pull(species_pair),
                       SpeciesPairClade = c("Cross", "Cross", "Angio", "Gymno", "Cross", "Cross", "Gymno", "Cross",
                       "Angio", "Cross", "Cross", "Cross", "Angio", "Gymno", "Cross"))


weighted_max_cliques <-  left_join(weighted_max_cliques , speciesPairs, by = join_by("species_pair" == "SpeciesPair" ))

weighted_max_cliques_filterable <- weighted_max_cliques%>% 
  mutate(Clade = paste0(SpeciesPairClade, "-sum" ))%>% 
  group_by(cliqueID)%>% 
  mutate(OriginalCliqueSize = n())%>%  
  mutate(OriginalCliqueSize = factor(OriginalCliqueSize))%>% 
  ungroup() %>% 
  pivot_wider(names_from = SpeciesPairClade, values_from = SpeciesPairClade)%>%
  pivot_wider(names_from = Clade, values_from = MaxpValNegLog10)

weighted_max_cliques_filterable <- replace_na(weighted_max_cliques_filterable, list(`Angio-sum` = 0, `Gymno-sum` = 0, `Cross-sum` = 0))

# The sum of the various pairs within a clique
weighted_max_cliques_filterable <- weighted_max_cliques_filterable %>% 
  group_by(cliqueID) %>% 
  mutate(AngioSum = sum(`Angio-sum`)) %>% 
  mutate(GymnoSum = sum(`Gymno-sum`)) %>% 
  mutate(CrossSum = sum(`Cross-sum`))


# Create columns with binary input for the various groups of pairs
weighted_max_cliques_filterable$Angio[is.na(weighted_max_cliques_filterable$Angio)] <- 0
weighted_max_cliques_filterable$Angio[weighted_max_cliques_filterable$Angio != 0] <-1
weighted_max_cliques_filterable$Gymno[is.na(weighted_max_cliques_filterable$Gymno)] <-0
weighted_max_cliques_filterable$Gymno[weighted_max_cliques_filterable$Gymno != 0] <-1
weighted_max_cliques_filterable$Cross[is.na(weighted_max_cliques_filterable$Cross)] <-0
weighted_max_cliques_filterable$Cross[weighted_max_cliques_filterable$Cross != 0] <-1

weighted_max_cliques_filterable <- weighted_max_cliques_filterable %>% 
  ungroup() %>% 
  mutate_at(c("Angio", "Gymno", "Cross"), as.numeric) %>% 
  mutate(UGeneSpecies1 = paste0(weighted_max_cliques_filterable$Species1, "-", weighted_max_cliques_filterable$GeneSpecies1),
         UGeneSpecies2 = paste0(weighted_max_cliques_filterable$Species2, "-", weighted_max_cliques_filterable$GeneSpecies2), .before = species_pair)

# ------------- SAVING POINT -----------------
# At this point the data set is ready to be used for identification of cliques.
save(weighted_max_cliques_filterable, file = "Data/DATA/clique_genes_filterable.RData")
```


**Step 3: identifying conserved and differentiated genes**


```{r}

SELECTION_METHOD <- "PVAL" # options: "SUM" = clique sum OR "PVAL" = p-value
TYPE_OF_GENE_SET <- "d" # options: "c" = conserved, "d" = differentiated

# --------------- PARAMETERS ---------------

P_VAL <- 0.1 # P_VAL < x
N_SIG_TOTAL <- 10 # NUMBER OF CROSSED PAIRS REQUIRED FOR CONSERVED CLIQUES (>= X) X/15
N_SIG_CLADE <- 3 # NUMBER OF CLADE PAIRS REQUIRED FOR CONSERVED CLIQUES (== X)  X/3
N_SIG_CROSS_D <- 3 # NUMBER OF CROSSED PAIRS PERMITTED FOR DIFFERENTIATED CLIQUES (<= X) X/9


```


```{r}

load("Data/DATA/clique_genes_filterable.RData")
# load("Data/DATA/ones_and_zeros_all.RData")

```



```{r eval=FALSE, include=FALSE}

# Checking the distribution of clique sums

library(ggplot2)

ggplot(weighted_max_cliques_filterable, aes(x = NegLog10CliqueSum)) + geom_density(color="red", linewidth = 1) + geom_vline(xintercept = 15, linewidth = 1)

```



```{r}
# ----------------CONSERVED GENES------------------------

if(TYPE_OF_GENE_SET == "c"){
  print(TYPE_OF_GENE_SET)
  
  if(SELECTION_METHOD == "PVAL"){
    print(SELECTION_METHOD)
    
    # conserved_genes <- weighted_max_cliques_filterable %>%
    #   filter(Max.p.Val < P_VAL) %>%
    #   group_by(cliqueID) %>%
    #   filter(sum(Angio) == N_SIG_CLADE & sum(Gymno) == N_SIG_CLADE & sum(Cross) >= N_SIG_CROSS_C) %>% 
    #   filter(OriginalCliqueSize == 15)
    
    genes <- weighted_max_cliques_filterable %>%
      select(c(1:7,9,10,12)) %>% 
      filter(Max.p.Val < P_VAL) 
    
    rm(weighted_max_cliques_filterable)
    
    genes <- genes %>% 
      filter(OriginalCliqueSize == 15) %>%  
      group_by(cliqueID) %>%
      mutate(size = n()) %>%
      filter(size >= N_SIG_TOTAL) %>%  # We require n significant pairs per clique
      ungroup()

    # Ensuring that all six species are present
    s1 <- genes %>% 
      select(UGeneSpecies1, Species1,GeneSpecies1,OrthoGroup, cliqueID) %>% 
      rename("Genes" = GeneSpecies1) %>% 
      rename("UGenes" = UGeneSpecies1)%>% 
      rename("Species" = Species1)
    
    s2 <- genes %>% 
      select(UGeneSpecies2, Species2,GeneSpecies2,OrthoGroup, cliqueID) %>% 
      rename("Genes" = GeneSpecies2) %>% 
      rename("UGenes" = UGeneSpecies2) %>% 
      rename("Species" = Species2)
    
    combined <- rbind(s1,s2)
    
    combined <- combined %>% 
      group_by(cliqueID) %>%
      distinct(UGenes, .keep_all = T) %>%
      mutate(size = n()) %>%
      filter(size == 6) %>% # Ensuring all species are present (despite not all pairs being present)
      ungroup() 
    
    # test <- conserved_genes %>%
    #   # filter(size == 11) %>%
    #   filter(cliqueID == "OG0000039_1172")
    
    genes <- genes %>% 
      filter(cliqueID %in% combined$cliqueID)
      
    
    print(length(unique(genes$OrthoGroup)))
    print(length(unique(genes$cliqueID)))
    
    file_name <- paste0("Data/DATA/conserved_genes_", SELECTION_METHOD, "_", P_VAL, ".RData") 
  } 
  
}




# ------------------DIFFERENTIATED GENES----------------------

if(TYPE_OF_GENE_SET == "d"){
  print(TYPE_OF_GENE_SET)
  
  if(SELECTION_METHOD == "PVAL"){
    print(SELECTION_METHOD)
    
    genes <- weighted_max_cliques_filterable %>% 
      filter(Max.p.Val < P_VAL) %>% 
      group_by(cliqueID) %>% 
      filter(sum(Angio) == N_SIG_CLADE & sum(Gymno) == N_SIG_CLADE & sum(Cross) <= N_SIG_CROSS_D) %>% 
      ungroup() %>% 
      filter(OriginalCliqueSize == 15) %>% 
      select(c(1:7,9,10,12))
    
    print(length(unique(genes$OrthoGroup)))
    print(length(unique(genes$cliqueID)))
    
    file_name <- paste0("Data/DATA/differentiated_genes_", SELECTION_METHOD, "_", P_VAL,"_XPAIRS_",N_SIG_CROSS_D ,".RData")
 
  } 

  
}

    save(genes, file = file_name)

```



# Differentiated genes: removing  conserved genes

1. Load the conserved cliques
2. Use the UGenes to filter the differentiated clique effectively removing genes that are involved in a conserved clique


```{r}

load("Data/DATA/conserved_genes_PVAL_0.1.RData") # PVAL

cons_genes <- genes
length(unique(cons_genes$OrthoGroup))
length(unique(cons_genes$cliqueID))

load("Data/DATA/differentiated_genes_PVAL_0.1_XPAIRS_3.RData") # PVAL

diff_genes <- genes
length(unique(diff_genes$OrthoGroup))
length(unique(diff_genes$cliqueID))

s1 <- diff_genes %>% 
  select(OrthoGroup, cliqueID, Species1, GeneSpecies1, UGeneSpecies1) %>% 
  rename("UGenes" = UGeneSpecies1, "Genes" = GeneSpecies1 ,"Species" = Species1)

s2 <- diff_genes %>% 
  select(OrthoGroup, cliqueID, Species2, GeneSpecies2, UGeneSpecies2) %>% 
  rename("UGenes" = UGeneSpecies2, "Genes" = GeneSpecies2,"Species" = Species2) 

combined_diff <- rbind(s1, s2)
combined_diff <- combined_diff %>% 
  ungroup() 

length(unique(combined_diff$cliqueID))

combined_diff <- combined_diff %>% 
  filter(!(UGenes %in% cons_genes$UGeneSpecies1)) %>% 
  filter(!(UGenes %in% cons_genes$UGeneSpecies2)) %>% 
  group_by(cliqueID) %>%   
  distinct(Species, .keep_all = T) %>% 
  mutate(sum = n()) %>% 
  filter(sum == 6) # require all six species, but not all pairs


length(unique(combined_diff$OrthoGroup))
length(unique(combined_diff$cliqueID))

# Filter the original file with the differentiated clique genes to retain same df format as the conserved genes

differentiated_genes <- diff_genes %>% 
  filter(cliqueID %in% combined_diff$cliqueID)

length(unique(differentiated_genes$OrthoGroup))
length(unique(differentiated_genes$cliqueID))

save(differentiated_genes, file = "Data/DATA/differentiated_genes_PVAL_0.1_filtered.RData")

```

# Formatting - longer data frames

For plotting and creating excel files.

```{r}

# GENES CONSERVED ACROSS CLADES

load("Data/DATA/conserved_genes_PVAL_0.1.RData") # conserved across all species

caas <- genes %>% 
  ungroup()

s1 <- caas %>% 
  select(UGeneSpecies1, Species1,GeneSpecies1,OrthoGroup, cliqueID, Max.p.Val) %>% 
  rename("Genes" = GeneSpecies1) %>% 
  rename("UGenes" = UGeneSpecies1)%>% 
  rename("Species" = Species1)

s2 <- caas %>% 
  select(UGeneSpecies2, Species2,GeneSpecies2,OrthoGroup, cliqueID, Max.p.Val) %>% 
  rename("Genes" = GeneSpecies2) %>% 
  rename("UGenes" = UGeneSpecies2) %>% 
  rename("Species" = Species2)

df <- rbind(s1,s2)  

# length(unique(df$cliqueID))  
# length(unique(df$OrthoGroup)) 

df <- df %>%
  group_by(cliqueID) %>%
  distinct(UGenes, .keep_all = T) %>%
  mutate(MeanCliqueSum = mean(Max.p.Val)) %>%
  ungroup() %>%
  select(-c(UGenes, Max.p.Val))

head(df)
# length(unique(df$cliqueID))  
# length(unique(df$OrthoGroup))  

caas_wider <- df %>% 
  pivot_wider(names_from = Species, values_from = Genes) %>% 
  select(OrthoGroup, cliqueID, MeanCliqueSum,Asp, Birch, Cher, Nor, Scots, Lodge)

head(caas_wider)

caas_wider_unique <- caas_wider %>% 
  group_by(OrthoGroup) %>% 
  arrange(MeanCliqueSum, .by_group = T) %>% 
  distinct(Asp,  .keep_all = T) %>% 
  distinct(Birch, .keep_all = T) %>% 
  distinct(Cher, .keep_all = T) %>% 
  distinct(Nor, .keep_all = T) %>% 
  distinct(Scots, .keep_all = T) %>% 
  distinct(Lodge, .keep_all = T) 


save(caas_wider, file = "Data/DATA/conserved_genes_wide.RData")
save(caas_wider_unique, file = "Data/DATA/conserved_genes_unique_wide.RData")


```

```{r}

#  GENES DIFFERENTIATED BETWEEN CLADES

load("Data/DATA/differentiated_genes_PVAL_0.1_filtered.RData")

dbc <- differentiated_genes %>%
  ungroup()
  
s1 <- dbc %>% 
  select(UGeneSpecies1, Species1,GeneSpecies1,OrthoGroup, cliqueID, Max.p.Val) %>% 
  rename("Genes" = GeneSpecies1) %>% 
  rename("UGenes" = UGeneSpecies1)%>% 
  rename("Species" = Species1)

s2 <- dbc %>% 
  select(UGeneSpecies2, Species2,GeneSpecies2,OrthoGroup, cliqueID, Max.p.Val) %>% 
  rename("Genes" = GeneSpecies2) %>% 
  rename("UGenes" = UGeneSpecies2) %>% 
  rename("Species" = Species2)

df <- rbind(s1,s2)  

# length(unique(df$cliqueID))  
# length(unique(df$OrthoGroup)) 

df <- df %>%
  group_by(cliqueID) %>%
  distinct(UGenes, .keep_all = T) %>%
  mutate(MeanCliqueSum = mean(Max.p.Val)) %>%
  ungroup() %>%
  select(-c(UGenes, Max.p.Val))

length(unique(df$cliqueID))
length(unique(df$OrthoGroup))

dbc_wider <- df %>% 
  pivot_wider(names_from = Species, values_from = Genes) %>% 
  select(OrthoGroup, cliqueID, MeanCliqueSum, Asp, Birch, Cher, Nor, Scots, Lodge)


dbc_wider_unique <- dbc_wider %>% 
  group_by(OrthoGroup) %>% 
  arrange(MeanCliqueSum, .by_group = T) %>% 
  distinct(Asp,  .keep_all = T) %>% 
  distinct(Birch, .keep_all = T) %>% 
  distinct(Cher, .keep_all = T) %>% 
  distinct(Nor, .keep_all = T) %>% 
  distinct(Scots, .keep_all = T) %>% 
  distinct(Lodge, .keep_all = T) 

save(dbc_wider, file = "Data/DATA/differentiated_genes_wide.RData")
save(dbc_wider_unique, file = "Data/DATA/differentiated_genes_unique_wide.RData")

```



