---
title: "Comparative sample heatmaps"
author: "Ellen Dimmen Chapple"
date: "2023-08-25"
output: 
  html_document:
    code_folding: hide
    toc: true
    number_sections: true
    toc_float: true
    theme: paper
editor_options:
  chunk_output_type: inline
---

```{r Packages, message=FALSE, warning=FALSE}

library(tidyverse)
library(ggplot2)
library(ComplexHeatmap)
library(cowplot)
library(gdata)
library(circlize)
library(patchwork)
library(RColorBrewer)
library(reshape2)

dir <- "../"
VER <- "V5" # Run version of ComPlEx
species_list <- c("Lodge", "Asp", "Nor","Scots","Birch","Cher")
combo <- data.frame(t(combn(species_list, 2)))

```


```{r file names}

# List of file names.

file_list_1 <- c()
for (i in 1:nrow(combo)){
  
  s1 <- combo[i, "X1"]
  s2 <- combo[i, "X2"]
  
  file_name <- paste0(dir,"Data/comparisonFiles/comparison-",s1,"_",s2,"-pearsonMR0.03no-table-",VER,".RData")
  file_list_1 <- append(file_list_1, file_name, after = length(file_list_1))
    
}

file_list_2 <- c()
for (i in species_list){
  
  expr_name <- paste0(dir,i,"Wood/",i,"Wood_transcriptomics.txt.gz")
  file_list_2 <- append(file_list_2, expr_name, after = length(file_list_2))
    
}


```


## Sample comparison heatmaps (fig 2C)

Creates 15 individual heatmaps showing level of correlation between species pairs across samples based on the expression levels of co-expressologs. Six additional heatmaps are created along the diagonal of the final figure (Fig 2C) that are each of the six species correlated against themselves using all expressed genes.

```{r corr matrix, fig.height=7, fig.width=7}

pval <- 0.1

# ---- Create correlation matrices for each species pair ----
for (x in file_list_1){
  
  load(x) 
  
  co_expressologs <- comparison_table%>%
    select(Species1, Species2,OrthoGroup, Max.p.val) %>%
    mutate_at("Max.p.val", as.numeric) %>%
    filter(Max.p.val < pval)  %>% 
    ungroup() %>%
    arrange(Max.p.val) %>% 
    distinct(Species1,  .keep_all = T) %>%
    distinct( Species2, .keep_all = T) # Each gene can only appear in one co-expressolog
  
  # SPECIES 1
  species_1 <- sapply(strsplit(x, "_"), "[",1) 
  species_1 <- sapply(strsplit(species_1, "-"), "[",2) 

  expr_data <- read.delim(paste0(dir,species_1,"Wood/", species_1,"Wood_transcriptomics.txt.gz")) 
  s1_expression <- co_expressologs %>%
    left_join(expr_data, join_by(Species1 == Genes)) %>%
    select(-c("Species2", "Max.p.val", "OrthoGroup")) %>%
    as_tibble() %>%
    column_to_rownames(var = "Species1") %>%
    select(contains("1."))
  
  s1_expression[is.na(s1_expression)] <- 0
  
  # SPECIES 2
  species_2 <- sapply(strsplit(x, "_"), "[",2)
  species_2 <- sapply(strsplit(species_2, "-"), "[",1) 
  
  expr_data <- read.delim(paste0(dir,species_2,"Wood/", species_2,"Wood_transcriptomics.txt.gz")) 
  
  s2_expression <- co_expressologs %>%
    left_join(expr_data, join_by(Species2 == Genes)) %>%
    select(-c("Species1", "Max.p.val", "OrthoGroup")) %>%
    as_tibble() %>%
    column_to_rownames(var = "Species2") %>%
    select(contains("1."))
  
  s2_expression[is.na(s2_expression)] <- 0
  
  
  # Correlation matrix
  corr_matrix <- round(cor(s1_expression,s2_expression, method = "pearson"), 4) 
  
  new_name <- as.character(paste0("corr_",species_1, "-",species_2))
  mv("corr_matrix", new_name) 
}

# Transposing matrices
to_transpose <- c("corr_Asp-Scots", "corr_Asp-Nor", "corr_Asp-Cher", "corr_Asp-Birch", "corr_Birch-Cher", "corr_Nor-Scots")
for(x in to_transpose){

    species_1 <- sapply(strsplit(x, "_"), "[",2)
    species_1 <- sapply(strsplit(species_1, "-"), "[",1)
    species_2 <- sapply(strsplit(x, "-"), "[",2) 

    corr_t <- t(get(x))
    new_name <- as.character(paste0("corr_",species_2, "-",species_1))
    mv("corr_t", new_name) 
    rm(list = x)
    
}

## Self-correlation matrices (using all expressed genes)
load(paste0(dir,"DATA/all_expressed_genes.RData"))

co_expressologs_all <- expr_genes %>% 
  filter(Max.p.Val < 0.1)

for(species in species_list){

  expr_data <- read.delim(paste0(dir,species,"Wood/", species,"Wood_transcriptomics.txt.gz")) 

  co_expressologs_one_species <- co_expressologs_all %>% 
    filter(Species1 ==  species | Species2 == species)
  
  col_1 <-co_expressologs_one_species %>% 
    select(Species1, GeneSpecies1) %>% 
    rename(Species = "Species1", Genes = "GeneSpecies1") %>% 
    filter(Species == species)
  
  col_2 <-co_expressologs_one_species %>% 
    select(Species2, GeneSpecies2) %>% 
    rename(Species = "Species2", Genes = "GeneSpecies2")%>% 
    filter(Species == species)
  
  if(nrow(col_1) == 0) {
    combined_df <- col_2
  }
  
  if(nrow(col_2) == 0) {
    combined_df <- col_1
  }

  if(nrow(col_2) > 0 & nrow(col_1) > 0){
    combined_df <- rbind(col_1, col_2)
  }
  
  combined_df <- combined_df %>% 
    distinct(Genes, .keep_all = T)
    
  
  expression <- expr_data %>%
    filter(Genes %in% combined_df$Genes) %>% 
    select(contains("1."))

  corr_matrix <- round(cor(expression[,-1], method = "pearson"), 4)
  new_name <- as.character(paste0("corr_",species, "-",species))
  mv("corr_matrix", new_name) # rename the heatmaps for plotting in grid.

}


# ---- Individual heatmaps ----

corr_list <- c("corr_Cher-Asp", "corr_Birch-Asp", "corr_Lodge-Asp", "corr_Nor-Asp", "corr_Scots-Asp", "corr_Cher-Birch", "corr_Nor-Birch", "corr_Nor-Cher", "corr_Scots-Birch", "corr_Scots-Cher", "corr_Scots-Nor", "corr_Lodge-Birch", "corr_Lodge-Cher", "corr_Lodge-Nor", "corr_Lodge-Scots", "corr_Scots-Scots", "corr_Nor-Nor", "corr_Lodge-Lodge", "corr_Asp-Asp", "corr_Birch-Birch", "corr_Cher-Cher"
)

col_function_1 <- rev(colorRampPalette(brewer.pal(n = 8, "YlOrBr"))(20))
for(matrix in corr_list){
  
  species_1 <- sapply(strsplit(matrix, "_"), "[",2)
  species_1 <- sapply(strsplit(species_1, "-"), "[",1)
  species_2 <- sapply(strsplit(matrix, "-"), "[",2)
  
  lowest <- min(get(matrix))
  highest <- max(get(matrix))
  between <- median(get(matrix))
  
  
  hm <-  Heatmap(get(matrix),show_column_names = FALSE,
                 show_row_names = FALSE,
                 width = ncol(get(matrix))*unit(1, "mm"),
                 height = nrow(get(matrix))*unit(1, "mm"),
                 row_order = 1:nrow(get(matrix)),
                 column_order = 1:ncol(get(matrix)), col = col_function_1, show_heatmap_legend = F
                 #         ,heatmap_width = unit(3, "cm"),
                 # heatmap_height = unit(3, "cm")
  )
  
  new_name <- as.character(paste0(species_1, "_",species_2))
  mv("hm", new_name) # rename the heatmaps for plotting in grid.

}

# ---- Convert Compexheatmap objects to ggplot objects

ba <- grid.grabExpr(print(Birch_Asp))
ca <- grid.grabExpr(print(Cher_Asp)) 
cb <- grid.grabExpr(print(Cher_Birch)) 
na <- grid.grabExpr(print(Nor_Asp)) 
nb <- grid.grabExpr(print(Nor_Birch))
nc <- grid.grabExpr(print(Nor_Cher))
sa <- grid.grabExpr(print(Scots_Asp)) #
sb <- grid.grabExpr(print(Scots_Birch))
sc <- grid.grabExpr(print(Scots_Cher))
sn <- grid.grabExpr(print(Scots_Nor))
la <- grid.grabExpr(print(Lodge_Asp))
lb <- grid.grabExpr(print(Lodge_Birch))
lc <- grid.grabExpr(print(Lodge_Cher))
ln <- grid.grabExpr(print(Lodge_Nor))
ls <- grid.grabExpr(print(Lodge_Scots))
# Diag
aa <- grid.grabExpr(print(Asp_Asp))
bb <- grid.grabExpr(print(Birch_Birch))
cc <- grid.grabExpr(print(Cher_Cher))
nn <- grid.grabExpr(print(Nor_Nor))
ss <- grid.grabExpr(print(Scots_Scots))
ll <- grid.grabExpr(print(Lodge_Lodge))

# Plot in correct order
hm <-plot_grid(aa,NULL, NULL, NULL, NULL,NULL,
               ba, bb, NULL, NULL,NULL, NULL,
               ca, cb, cc, NULL, NULL, NULL,
               na, nb, nc, nn, NULL, NULL,
               sa, sb,sc, sn, ss, NULL,
               la, lb, lc, ln, ls, ll, nrow = 6, ncol = 6)

hm
 # pdf(paste0(dir,"Figures/sampleCompHM.pdf"), height=7, width = 7)
 # hm
 # dev.off()
```


## Density plots of sample correlation values (supplementary fig 2B)

```{r fig.height=10, fig.width=5}

df_for_density <- c()
for(matrix in corr_list){
  
  pair <- sapply(strsplit(matrix, "_"), "[",2)
  correlation_vect <- c(get(matrix))
  
  df <- data.frame(
    pair = rep(pair, length(correlation_vect)),
    correlation_values = correlation_vect)
  
  df_for_density <- rbind(df_for_density, df)
  

}


angio_pairs <- data.frame(pair = c("Cher-Asp" ,  "Birch-Asp", "Cher-Birch" ))
gymno_pairs <- data.frame(pair = c(  "Lodge-Nor" , "Lodge-Scots", "Scots-Nor"))
self <- data.frame(pair = c(   "Scots-Scots" ,"Nor-Nor"   ,  "Lodge-Lodge", "Asp-Asp" ,    "Birch-Birch", "Cher-Cher"  ))


all_non_self <- df_for_density %>%
  filter( !(pair %in% self$pair))

all_non_self[all_non_self == "Lodge-Scots"] <- "P.contorta - P.sylvestris"
all_non_self[all_non_self == "Lodge-Nor"] <- "P.contorta - P.abies"
all_non_self[all_non_self == "Scots-Nor"] <- "P.sylvestris - P.abies"
all_non_self[all_non_self == "Lodge-Cher"] <- "P.contorta - P.avium"
all_non_self[all_non_self == "Lodge-Birch"] <- "P.contorta - B.pendula"
all_non_self[all_non_self == "Scots-Cher"] <- "P.sylvestris - P.avium"
all_non_self[all_non_self == "Scots-Birch"] <- "P.sylvestris - B.pendula"
all_non_self[all_non_self == "Nor-Cher"] <- "P.abies - P.avium"
all_non_self[all_non_self == "Nor-Birch"] <- "P.abies - B.pendula"
all_non_self[all_non_self == "Scots-Asp"] <- "P.sylvestris - P.tremula"
all_non_self[all_non_self == "Nor-Asp"] <- "P.abies - P.tremula"
all_non_self[all_non_self == "Lodge-Asp"] <- "P.contorta - P.tremula"
all_non_self[all_non_self == "Cher-Birch"] <- "P.avium - B.pendula"
all_non_self[all_non_self == "Birch-Asp"] <- "B.pendula - P.tremula"
all_non_self[all_non_self == "Cher-Asp"] <- "P.avium - P.tremula"

speciesPairs <- tibble(SpeciesPair = all_non_self %>% distinct(pair) %>% pull(pair),
                       SpeciesPairClade = c("Angiosperm", "Angiosperm", "Crossed", "Crossed", "Crossed", "Angiosperm", "Crossed", "Crossed",
                       "Crossed", "Crossed", "Gymnosperm", "Crossed", "Crossed", "Gymnosperm", "Gymnosperm"))


all_non_self <-  left_join(all_non_self , speciesPairs, by = join_by("pair" == "SpeciesPair" ))

all_non_self <- all_non_self%>%
  mutate(pair = fct_relevel(pair,  "P.avium - P.tremula", "B.pendula - P.tremula", "P.avium - B.pendula",  "P.contorta - P.tremula" ,  "P.abies - P.tremula",     "P.sylvestris - P.tremula",   "P.abies - B.pendula"  ,   "P.abies - P.avium","P.sylvestris - B.pendula", "P.sylvestris - P.avium"  , "P.contorta - B.pendula", "P.contorta - P.avium",  "P.sylvestris - P.abies"  ,"P.contorta - P.abies", "P.contorta - P.sylvestris"))


angio_density <- all_non_self %>% 
  filter(SpeciesPairClade == "Angiosperm") %>% 
  group_by(pair) 

gymno_density <- all_non_self %>% 
  filter(SpeciesPairClade == "Gymnosperm") %>% 
  group_by(pair)

crossed_density <- all_non_self %>% 
  filter(SpeciesPairClade == "Crossed") %>% 
  group_by(pair)

# Line plot - lines
title_size <- 20
axis_title_size <- 20
axis_text_size <- 20

p1 <-  ggplot(angio_density, aes(x = correlation_values, groups = pair)) + 
  geom_density(linewidth = 2, colour = "#FC8D62") + 
  labs(title = "Dicot pairs",x = "", y = "Density") +  
  ylim(0,4)+ xlim(0,1) +  
  theme_bw()+ 
  theme(plot.title = element_text(hjust = 1, size = title_size), axis.title = element_text(size = axis_title_size, vjust = 2.5), axis.text =  element_text(size = axis_title_size))


p2 <-ggplot(gymno_density, aes(x = correlation_values, groups = pair)) + 
  geom_density(linewidth = 2, colour = "#66C2A5") + 
  labs(title = "Conifer pairs",x = "", y = "Density") +  
  ylim(0,4)+ xlim(0,1) + 
  theme_bw()+ 
  theme(plot.title = element_text(hjust = 1, size = title_size), axis.title = element_text(size = axis_title_size, vjust = 2.5), axis.text =  element_text(size = axis_title_size))



p3 <-ggplot(crossed_density, aes(x = correlation_values, groups = pair)) +
  geom_density(linewidth = 2) + labs( title = "Inter-lineage pairs",x = "Correlation", y = "Density") +  
  ylim(0,4)+ xlim(0,1) + 
  theme_bw()+
  theme(plot.title = element_text(hjust = 1, size = title_size), axis.title = element_text(size = axis_title_size, vjust = 2.5), axis.text =  element_text(size = axis_title_size))

p1/p2/p3

# pdf(paste0(dir,"Figures/density_plots.pdf"), height =10, width = 5)
# p1/p2/p3
# dev.off()

```


























