---
title: "R ComPlEx"
author: "Torgeir R Hvidsten"
date: '`r format(Sys.time(), "%d.%m.%Y")`'
output:
  html_document:
    theme: yeti
    code_folding: hide
    number_sections: false
editor_options: 
  chunk_output_type: console
---

```{r libraries, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = FALSE)

library(tidyverse)
theme_set(theme_classic())
theme_update(plot.title = element_text(face="bold"))

library(gplots)
library("RColorBrewer")
library(DT)
library(cowplot)
library(matrixStats)
```

### Set up species and input files

```{r setup}

# Species name
species1_name <- "Asp"
species2_name <- "Nor"

# Expression files
species1_expr_file <- "../../AspWood/AspWood_transcriptomics.txt.gz"
species2_expr_file <- "../../NorWood/NorWood_transcriptomics.txt.gz"

# Ortholog groups
ortholog_group_file <- "../../DATA/OrthoGroups/Orthogroups_20240823_clean.tsv.gz"

cat("ComPlEX:", species1_name, species2_name, "\n")

```

### Read in ortholog groups and expression data

```{r input, warning=FALSE, error=FALSE, message=FALSE}

# Read in ortholog groups
# =======================

# Reading in and parsing the orthogroup file takes time. Parsed results are saved
# and used if they exists.
ortholog_group_RData <- paste0("RData/orthologs-", species1_name, "-", species2_name, ".RData")

if (!file.exists(ortholog_group_RData)) {
  
  # Asp Nor
  ortho <- read_tsv(ortholog_group_file, show_col_types = FALSE) %>%
    dplyr::rename(Species1 = Populus_tremula, Species2 = Picea_abies) %>%
    select(Species1, Species2, OrthoGroup) %>%
    filter(Species1 != "", Species2 != "") %>%
    separate_rows(Species1, sep = ", ", convert = FALSE) %>%
    separate_rows(Species2, sep = ", ", convert = FALSE) %>%
    group_by(Species1, Species2, OrthoGroup) %>%
    dplyr::slice(1) %>% 
    ungroup()
  
  # Add annotations from arabidopsis
  symbols <- read_tsv("../../DATA/Annotations/gene_aliases_20140331.txt", show_col_types = FALSE) %>%
    dplyr::rename(Arab = locus_name, Symbol = symbol, Name = full_name)
  
  annot <- read_tsv(ortholog_group_file, show_col_types = FALSE) %>%
    dplyr::rename(Arab = Arabidopsis_thaliana) %>%
    select(Arab, OrthoGroup) %>%
    filter(Arab != "") %>%
    separate_rows(Arab, sep = ", ", convert = FALSE) %>%
    left_join(symbols, by = "Arab") %>%
    group_by(OrthoGroup) %>%
    summarise(Arab = paste0(unique(Arab), collapse = "; "), 
           Symbol = paste0(unique(Symbol), collapse = "; "), 
           Name = paste0(unique(Name), collapse = "; ")) %>%
    mutate(Symbol = gsub("NA", "", Symbol),
           Name = gsub("NA", "", Name))

  save(ortho, annot, file = ortholog_group_RData)
} else {
  load(file = ortholog_group_RData)
}

# Read in expression data
# =======================

species1_expr <- read.delim(species1_expr_file, sep = "\t", header = TRUE)

species2_expr <- read.delim(species2_expr_file, sep = "\t", header = TRUE)

# Filter
# ======

cat (length(unique(ortho$OrthoGroup)), " ortholog groups containing:\n",
     " ", length(unique(ortho$Species1)), " ", species1_name, " genes\n",
     " ", length(unique(ortho$Species2)), " ", species2_name, " genes\n\n",
     length(unique(species1_expr$Genes)), " expressed ", species1_name, " genes\n",
     length(unique(species2_expr$Genes)), " expressed ", species2_name, " genes\n",
     sep = "")

ortho <- ortho %>%
  filter(Species1 %in% species1_expr$Genes & Species2 %in% species2_expr$Genes)

species1_expr <- species1_expr[species1_expr$Genes %in% ortho$Species1,]
species2_expr <- species2_expr[species2_expr$Genes %in% ortho$Species2,]

cat ("After filtering on expressed genes with ortholog:\n",
     " ", length(unique(ortho$OrthoGroup)), " ortholog groups containing: \n",
     "  ", length(unique(ortho$Species1)), " ", species1_name, " genes\n",
     "  ", length(unique(ortho$Species2)), " ", species2_name, " genes\n",
     sep = "")

```

### Compute co-expression networks

Several parameters:

- Correlation: Pearson or Spearman
- Normalization: CLR or MR
- Network density threshold

```{r coexpression}
cor_method <- "pearson" # pearson spearman
cor_sign <- "" # abs
norm_method <- "MR" # CLR MR
density_thr <- 0.03
randomize <- "no" # yes no

comparison_RData <- paste0("RData/comparison-", species1_name, "-", species2_name, "-", 
                           cor_sign, cor_method, norm_method, density_thr, randomize, ".RData")

if (!file.exists(comparison_RData)) {
  
  if (randomize == "yes") {
    species1_expr$Genes <- sample(species1_expr$Genes, nrow(species1_expr), FALSE)
    species2_expr$Genes <- sample(species2_expr$Genes, nrow(species2_expr), FALSE)
  }
  
  species1_net <- cor(t(species1_expr[,-1]), method = cor_method)
  dimnames(species1_net) <- list(species1_expr$Genes, species1_expr$Genes)
  
  species2_net <- cor(t(species2_expr[,-1]), method = cor_method)
  dimnames(species2_net) <- list(species2_expr$Genes, species2_expr$Genes)
  
  if (cor_sign == "abs") {
    species1_net <- abs(species1_net)
    species2_net <- abs(species2_net)
  }
  
  if (norm_method == "CLR") {
    #species1_net <- matrix(c(1,0.1,0.2,0.1,1,1,0.2,1,1), nrow = 3, byrow = TRUE)
    
    z <- scale(species1_net)
    z[z < 0] <- 0
    species1_net <- sqrt(t(z)**2 + z**2)
    
    z <- scale(species2_net)
    z[z < 0] <- 0
    species2_net <- sqrt(t(z)**2 + z**2)
    
  } else if (norm_method == "MR") {
    R <- t(apply(species1_net, 1, rank))
    species1_net <- sqrt(R * t(R))
    
    R <- t(apply(species2_net, 1, rank))
    species2_net <- sqrt(R * t(R))
  }
  
  diag(species1_net) <- 0
  diag(species2_net) <- 0
  
  R <- sort(species1_net[upper.tri(species1_net, diag = FALSE)], decreasing = TRUE)
  species1_thr <- R[round(density_thr*length(R))]
  plot(density(R), xlab = paste0(species1_name, " correlations"), main = "")
  
  R <- sort(species2_net[upper.tri(species2_net, diag = FALSE)], decreasing = TRUE)
  species2_thr <- R[round(density_thr*length(R))]
  plot(density(R), xlab = paste0(species2_name, " correlations"), main = "")
  
} else {
  load(file = comparison_RData)
}

cat("Correlation: ", cor_method, "\n",
    "Normalization: ", norm_method, "\n",
    "Density: ", density_thr, "\n",
    "Randomization: ", randomize, "\n\n",
    species1_name, " co-expr threshold: ", format(species1_thr, digits = 3) , "\n",
    species2_name, " co-expr threshold: ", format(species2_thr, digits = 3) , "\n",
    sep = "")


```

### Network comparison

For each ortholog pair, p-values for for the overlap of network neighborhoods are computed in both direction. 

```{r comparison, warning=FALSE, error=FALSE, message=FALSE}

if (!file.exists(comparison_RData)) {

  comparison <- ortho
  
  comparison$Species1.neigh <- c(NA)
  comparison$Species1.ortho.neigh <- c(NA)
  comparison$Species1.neigh.overlap <- c(NA)
  comparison$Species1.p.val <- c(NA)
  
  comparison$Species2.neigh <- c(NA)
  comparison$Species2.ortho.neigh <- c(NA)
  comparison$Species2.neigh.overlap <- c(NA)
  comparison$Species2.p.val <- c(NA)
  
  for (i in 1:nrow(ortho)) {
    
    if (i %% 100 == 0) {
      cat(i, "\n")
    }
    
    # Species 1 -> Species 2
    
    neigh <- species1_net[ortho$Species1[i],]
    neigh <- names(neigh[neigh >= species1_thr])
    
    ortho_neigh <- species2_net[ortho$Species2[i],]
    ortho_neigh <- names(ortho_neigh[ortho_neigh >= species2_thr])
    ortho_neigh <- ortho$Species1[ortho$Species2 %in% ortho_neigh]
    
    N <- nrow(species1_expr)
    m <- length(neigh)
    n <- N-m
    k <- length(unique(ortho_neigh))
    x <- length(unique(intersect(neigh, ortho_neigh)))
    p_val <- 1
    if (x > 1) {
      p_val <- phyper(x-1, m, n, k, lower.tail = FALSE)
    }
    
    comparison$Species1.neigh[i] <- m
    comparison$Species1.ortho.neigh[i] <- k
    comparison$Species1.neigh.overlap[i] <- x
    comparison$Species1.p.val[i] <- p_val
    
    # Species 2 -> Species 1
    
    neigh <- species2_net[ortho$Species2[i],]
    neigh <- names(neigh[neigh >= species2_thr])
    
    ortho_neigh <- species1_net[ortho$Species1[i],]
    ortho_neigh <- names(ortho_neigh[ortho_neigh >= species1_thr])
    ortho_neigh <- ortho$Species2[ortho$Species1 %in% ortho_neigh]
    
    N <- nrow(species2_expr)
    m <- length(neigh)
    n <- N-m
    k <- length(unique(ortho_neigh))
    x <- length(unique(intersect(neigh, ortho_neigh)))
    p_val <- 1
    if (x > 1) {
      p_val <- phyper(x-1, m, n, k, lower.tail = FALSE)
    }
    
    comparison$Species2.neigh[i] <- m
    comparison$Species2.ortho.neigh[i] <- k
    comparison$Species2.neigh.overlap[i] <- x
    comparison$Species2.p.val[i] <- p_val
  }
  
  save(comparison, species1_thr, species2_thr, file = comparison_RData)
}

# Filter orthologs not in the networks
comparison <- comparison %>%
  filter(Species1.neigh.overlap > 0 & Species2.neigh.overlap > 0)

# FDR correction
comparison$Species1.p.val <- p.adjust(comparison$Species1.p.val, method = "fdr")
comparison$Species2.p.val <- p.adjust(comparison$Species2.p.val, method = "fdr")

cat ("After filtering on gene pairs in the networks:\n",
     " ", length(unique(comparison$OrthoGroup)), " ortholog groups containing: \n",
     "  ", length(unique(comparison$Species1)), " ", species1_name, " genes\n",
     "  ", length(unique(comparison$Species2)), " ", species2_name, " genes\n",
     sep = "")

# Comparsion of p-values of orthologs: species 1 -> species 2 vs species 2 -> species 1
R <- cor.test(-log10(comparison$Species1.p.val), -log10(comparison$Species2.p.val))

data.frame(s1 = -log10(comparison$Species1.p.val),
           s2 = -log10(comparison$Species2.p.val)) %>%
  ggplot(aes(x = s1, y = s2)) +
  xlab(paste0(species1_name, " p-value (-log10)")) +
  ylab(paste0(species2_name, " p-value (-log10)")) +
  geom_point() + 
  geom_smooth(method=lm, formula = y ~ x, fill = "gainsboro") +
  ggtitle(paste0("Correlation = ", format(R$estimate, digits = 3)))

# Print some summary statistics
# =============================

# Gene pairs

cat("GENE PAIRS:\n",
    sum(comparison$Species1.p.val< 0.05), " conserved ", species1_name, " gene pairs (", 
    format(sum(comparison$Species1.p.val< 0.05)/nrow(comparison), digits = 3), ")\n", 
    sum(comparison$Species2.p.val< 0.05), " conserved ", species2_name, " gene pairs (", 
    format(sum(comparison$Species2.p.val< 0.05)/nrow(comparison), digits = 3), ")\n",
    sum(comparison$Species1.p.val< 0.05 & comparison$Species2.p.val< 0.05), " reciprocally conserved gene pairs (", 
    format(sum(comparison$Species1.p.val< 0.05 & comparison$Species2.p.val< 0.05)/nrow(comparison), digits = 3), ")\n", sep = "")

# Genes

comparison_species1 <- comparison %>%
  group_by(Species1) %>%
  arrange(Species1.p.val) %>%
  dplyr::slice(1)

comparison_species2 <- comparison %>%
  group_by(Species2) %>%
  arrange(Species2.p.val) %>%
  dplyr::slice(1)

comparison_species1_12 <- comparison %>%
  rowwise() %>%
  mutate(Max.p.val = max(Species1.p.val, Species2.p.val)) %>%
  group_by(Species1) %>%
  arrange(Max.p.val) %>%
  dplyr::slice(1)

comparison_species2_12 <- comparison %>%
  rowwise() %>%
  mutate(Max.p.val = max(Species1.p.val, Species2.p.val)) %>%
  group_by(Species2) %>%
  arrange(Max.p.val) %>%
  dplyr::slice(1)

cat("GENES:\n",
    sum(comparison_species1$Species1.p.val< 0.05), " conserved ", species1_name, " genes (", 
    format(sum(comparison_species1$Species1.p.val< 0.05)/nrow(comparison_species1), digits = 3), ")\n", 
    sum(comparison_species2$Species2.p.val< 0.05), " conserved ", species2_name, " genes (", 
    format(sum(comparison_species2$Species2.p.val< 0.05)/nrow(comparison_species2), digits = 3), ")\n",
    sum(comparison_species1_12$Max.p.val< 0.05), " reciprocally conserved ", species1_name, " genes (", 
    format(sum(comparison_species1_12$Max.p.val< 0.05 & 
               comparison_species1_12$Max.p.val< 0.05)/nrow(comparison_species1_12), digits = 3), ")\n",
    sum(comparison_species2_12$Max.p.val< 0.05), " reciprocally conserved ", species2_name, " genes (", 
    format(sum(comparison_species2_12$Max.p.val< 0.05 & 
               comparison_species2_12$Max.p.val< 0.05)/nrow(comparison_species2_12), digits = 3), ")\n",
    sep = "")

# Orthogroups

comparison_species1 <- comparison %>%
  group_by(OrthoGroup) %>%
  arrange(Species1.p.val) %>%
  dplyr::slice(1)

comparison_species2 <- comparison %>%
  group_by(OrthoGroup) %>%
  arrange(Species2.p.val) %>%
  dplyr::slice(1)

comparison_species12 <- comparison %>%
  rowwise() %>%
  mutate(Max.p.val = max(Species1.p.val, Species2.p.val)) %>%
  group_by(OrthoGroup) %>%
  arrange(Max.p.val) %>%
  dplyr::slice(1)

cat("ORTHOLOG GROUPS:\n",
    sum(comparison_species1$Species1.p.val< 0.05), " conserved ", species1_name, " orthogroups (", 
    format(sum(comparison_species1$Species1.p.val< 0.05)/nrow(comparison_species1), digits = 3), ")\n", 
    sum(comparison_species2$Species2.p.val< 0.05), " conserved ", species2_name, " orthogroups (", 
    format(sum(comparison_species2$Species2.p.val< 0.05)/nrow(comparison_species2), digits = 3), ")\n",
    sum(comparison_species12$Max.p.val< 0.05), " reciprocally conserved orthogroups (", 
    format(sum(comparison_species12$Max.p.val< 0.05 & 
               comparison_species12$Max.p.val< 0.05)/nrow(comparison_species12), digits = 3), ")\n", sep = "")

```

Table of all comparisons.

```{r, genepairs, warning=FALSE, error=FALSE, message=FALSE}

comparison_table <- comparison %>%
  rowwise() %>%
  mutate(Max.p.val = max(Species1.p.val, Species2.p.val)) %>%
  left_join(annot, by = "OrthoGroup") %>%
  select(-c("Species1.neigh", "Species1.ortho.neigh", "Species2.neigh", "Species2.ortho.neigh")) %>%
  arrange(Max.p.val)
  
comparison_table$Species1.p.val <- format(comparison_table$Species1.p.val, digits = 3, scientific = TRUE)
comparison_table$Species2.p.val <- format(comparison_table$Species2.p.val, digits = 3, scientific = TRUE)
comparison_table$Max.p.val <- format(comparison_table$Max.p.val, digits = 3, scientific = TRUE)

datatable(comparison_table, rownames = FALSE, filter = "top",
          options = list(
            columnDefs = list(list(className = 'dt-center', targets = "_all"))
            )
          )

save(comparison_table, file = paste0("RData/comparison-", species1_name, "-", species2_name,
                                     "-", cor_sign, cor_method, norm_method, density_thr, 
                                     randomize, "-", "table.RData"))

```

### Examples

```{r examples, warning=FALSE, error=FALSE, message=FALSE, fig.height = 15}

species1_expr_long <- species1_expr %>% 
  gather (Sample.names, Expression, -1) %>%
  separate(Sample.names, into = c("Trees", "Samples"), sep = "\\.", remove = FALSE) %>%
  mutate_at("Samples", as.numeric) %>%
  filter(Trees == "A1")

species2_expr_long <- species2_expr %>% 
  gather (Sample.names, Expression, -1) %>%
  separate(Sample.names, into = c("Trees", "Samples"), sep = "\\.", remove = FALSE) %>%
  mutate_at("Samples", as.numeric) %>%
  filter(Trees == "N1")

load("../../AspWood/gene.cluster.RData")
species1.d <- d
species1.gene.clusters <- gene.clusters
clusters <- unique(gene.clusters$Colors)

load("../../NorWood/gene.cluster.RData")
species2.d <- d

plots <- list()
for (c in clusters) {
  
  candidate_genes <- species1.gene.clusters %>%
    filter(Colors == c) %>%
    pull(Genes)

  candidate_gene <- comparison_table %>%
    ungroup() %>%
    filter(Species1 %in% candidate_genes) %>%
    dplyr::slice(1)
  
  id_species1 <- candidate_gene$Species1[1]
  name_species1 <- candidate_gene$Symbol[1]
    
  plots[[length(plots)+1]] <- species1_expr_long %>%
    filter(Genes == id_species1) %>%
    ggplot(aes(x = Samples, y = Expression, col = Genes)) +
    geom_line(size = 2) +
    ggtitle(paste0(id_species1, "(", name_species1, ")")) + ylab("Expression (VST)") +
    geom_vline(data = species1.d, mapping = aes(xintercept = xintercept), linetype="dashed", color = "gray68", linewidth=0.5) +
    theme(legend.position = "none") +
    scale_color_manual(values=as.vector(c))
  
  id_species2 <- candidate_gene$Species2[1]
  
  plots[[length(plots)+1]] <- species2_expr_long %>%
    filter(Genes == id_species2) %>%
    ggplot(aes(x = Samples, y = Expression, col = Genes)) +
    geom_line(size = 2) +
    ggtitle(id_species2) + ylab("Expression (VST)") +
    geom_vline(data = species1.d, mapping = aes(xintercept = xintercept), linetype="dashed", color = "gray68", linewidth=0.5) +
    theme(legend.position = "none") +
    scale_color_manual(values=as.vector(c))
    
}

plot_grid(plotlist = plots, ncol = 2)


```



